Anxiety, avoidance, and sequential evaluation

### 2019-08-01 manuscripts/manuscript/manuscript.tex
Updates from Overleaf

Though anxiety disorders differ in their particular symptomology, and in the content and situations that elicit symptoms, they all are similarly characterized by aberrations in the processing of and response to threat \citep{dsm5}. In particular, at least three symptoms manifest across the anxiety disorders. First, anxiety is associated with exaggerated threat appraisal, or a bias towards evaluating threat as disproportionately greater in likelihood and severity than is warranted \citep{ClarkBeck2011}. Second, anxiety is also associated with fear generalization, wherein the primary threat becomes associated with increasingly distal locations, events, and thoughts \citep{dymond2015}. Finally, anxiety is associated with persistent avoidance behavior, which often occurs well in advance of the materialization of actual threat \citep{Arnaudova2017}. Such inappropriate avoidance behaviors are especially harmful aspects of the disorders, both because they interfere with daily life and because they indirectly maintain anxiety by preventing learning from the non-occurrence of perceived threats. Though laboratory studies of decision making and learning in anxious populations have corroborated these clinical observations \citep{Harle2017, norbury2018, Aylward2019}, none offer an explanation as to the root of these symptoms.

@@

Although the return-maximizing assumption self-consistently defines optimal behavior, an agent need not be restricted to it \citep{symmonds2010}, and might in principle anticipate encountering danger under different (e.g., pathological) assumptions about the future. For example, an agent may expect to fail to take the correct protective actions in later states (i.e., to use a suboptimal $\pi(a \mid s)$), or may expect the world's future dynamics do not guarantee reliable avoidance even so (i.e., under stochastic or adversarial transition dynamics $P(s' \mid s,a)$). Consider an agent who has such pessimistic expectations about dangerous events at future steps. Note that assumptions of this sort, even if incorrect, can serve adaptive purposes. In general, pessimistic assumptions can help to ensure robustness and safety under uncertain or even adversarial scenarios \cite{Garcia2015}. Related work in RL shows how computing returns with pessimistic predictions can help to quantify variability in outcomes (i.e., to learn different points in the distribution of possible returns)\cite{bellemare2017}, which is one way to explicitly and tunably take account of risk tolerance. Indeed, it is common in machine learning theory to optimize outcomes under worst-case assumptions.

@@

The parameter $w$ controls the degree of pessimism. An optimistic agent ($w = 1$) expects in the future to act fully in accordance with its preferences, whereas a pessimistic agent ($w = 0$) expects to act contrary to its preferences. \Importantly, this belief is false under the model: that is, we assume, at each step, that the agent actually chooses by maximizing the values given by Equation~\ref{eq:pessimism}. But these values are always computed under the assumption that the agent will then not maximize at all later steps. (Another way to say this is that we assume choices at each step are optimal, but only under the assumption that later choices will not be.) Fig.~1c,d illustrates the the consequences of different levels of pessimism for valuation in an example gridworld. The value of threat propagates, with increasing distance across the state space, due to an increasing expectation that the agent may fail to avoid threat in the future.

@@

@@

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/02_appavo.png}}%
  }
  \par \textbf{Figure 2:} (A) The Balloon Analog Risk Task (BART)\citep{Lejuez2002,Maner2007, ramirez2015}. The risk of balloon burst (point loss) increases with each pump. The optimal policy (number of pumps) under increasingly pessimistic valuation is presented for low (B) and high risk (C) balloons. The optimistic agent ($w=1$) prefers a policy reflecting the true environmental risk. The moderate ($w=0.6$) and strongly ($w=0.2$) pessimistic agents cash-out earlier, as is observed in anxious individuals. (D/E) In the sleeping predator task, the risk of loss is constant but the cost of loss still increases as more rewards are gathered. The value of reward pursuit under increasingly pessimistic valuation is presented for scenarios with low (D) and high risk (E)  of predator awakening. The relative value of approach (vs. avoid) decreases with loss amount and threat level, and moreso under pessimistic assumptions. (Parameters: $\gamma = 1.0$)
\end{figure}

@@

So far, we have considered only the asymptotic preferences implied by our pessimistic value function, which we computed directly value iteration. But we can also consider the process of learning under this value function (e.g., by variants of Q-learning\citep{SuttonBarto2018} or DYNA\citep{sutton1991, russek_predictive_2017} using the $\beta$-pessimistic return). The dynamics of such learning may speak to the progression of symptoms.

@@

\begin{figure}[!b]
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/05_choice.png}}%
  }
  \par \textbf{Figure 4:} (A) The free choice premium paradigm\cite{Leotti2011,Leotti2014} with equal chance of outcome $R \in [1, -1]$. Non-anxious participants exhibit a preference for the free choice option (blue) despite it conferring no benefit over the fixed choice option (grey). (B) Pessimistic agents show an attenuated free choice bias. The fractional preference for the free choice option over the simulated experiment is shown for three populations of subjects with levels of pessimism $w$ (y-axis; each dot represents an individual simulated agent, and the smoothed density of free choice bias for each pessimism level is also shown.)
  (Parameters: Q-learning with $\gamma = 1.0$, and inverse temperature, $\beta$, increased from 1 to 10 over 100 episodes.)
\end{figure}

@@

For instance, one variant of the model is suggested by the observation that Eq.~1 is also computed in expectation over the anticipated future environmental dynamics $p(s' \mid s,a)$. Thus, pessimism can alternatively be encoded in this distribution: e.g., a false belief that the world's response to one's choices is unpredictable or adversarial. Because the Bellman equation for the return averages over this distribution in addition to the choice policy at each step, and because an unpredictable environment also reduces the efficacy of avoidance, either formulation can produce ultimately similar results in our simulations here. A third variant of our model arises from uncertainty about the current state $s$ of the environment. Although we have taken it as fully observed, if the world state is only partly known, then this distribution too must be averaged out in evaluating each action \citep{kaelbling1998}, and here also a pessimistic skew will propagate the expectation of danger and result in exaggerated avoidance. In summary, pessimistic resolution of several different varieties of uncertainty (e.g., about future action, environmental dynamics, or environmental state) could each produce similar symptoms for analogous reasons. However, from the perspective of cognitive theories of anxiety, these represent quite different maladaptive beliefs: a key difference that may be relevant in guiding treatment (especially cognitive psychotherapies aimed at ameliorating the false beliefs) of a host of anxiety disorders.

Particularly due to the way it encompasses several such variants, our account formalizes a longstanding range of theory on the role of control in anxiety. Central to many prominent cognitive theories of anxiety in the psychiatric literature is a perceived lack of control. For example, self-efficacy theory \citep{bandura1977} and the triple vulnerabilities model \citep{barlow2002} both posit that a reduced belief in the ability to effectively respond to future threat is involved in the genesis and maintenance of clinical anxiety. In contrast, and focused less on the self, the learned helplessness theories of anxiety \cite{alloy1990} claims clinical anxiety results from an uncertain belief in the controllability of the environment, such that future threat cannot be effectively mitigated or avoided. As we note above, the present model and analysis (though simulated above in terms of self-efficacy) can accommodate either variant, and shows how they relate to one another as members of a more general family of accounts. %Also, they are by no means mutually exclusive, though they might also reflect multiple distinct (and differently treatable) routes to similar symptomatology.

%An alternative variant of our proposal concerns the representation of states in the environment. In the present work, we considered only environments with fully observable states; in other words, there was no uncertainty regarding the state (e.g. grid world location) an agent currently occupied. In the real world, however, states must commonly be inferred from external and internal observations. This provides yet another source of uncertainty similar to the future policy and transition dynamics, which must be averaged over in evaluating actions \citep{kaelbling1998}. Thus another variant of our approach is to view anxiety as potentially also arising due to pessimistic biases in state inference\cite{Paulus2012}, which would again have similar effects on the resulting value function. For example, if an agent is erroneously biased toward believing herself more proximal to threat than is actually the case, then symptoms like exaggerated threat appraisal and avoidance should arise. Indeed, such a bias could result in the diffuse spread of threat across objects\citep{norbury2018} and location\citep{schulz2018}, thereby causing otherwise benign states to be perceived as threatening.

@@

It is important to note that the present model may not describe all anxiety disorders with equal accuracy. Indeed, our analysis of pessimistic sequential evaluation is, by definition, a model of prospective cognition. Thus, the present results are more likely to accurately describe the anxiety disorders which primarily involve aberrations in future-oriented cognitive processes, such as generalized and social anxiety disorders. Naturally, the present model can account neither for the compulsive behaviors of obsessive-compulsive disorder (OCD) nor the memory disturbances of post-traumatic stress disorder (PTSD). That said, recent clinical studies suggest that diminished perceived control is a vulnerability factor common to all anxiety disorders, including OCD and PTSD \citep{gallagher2014a, gallagher2014b}. Importantly, aspects of other psychiatric disorders that involve future-oriented misbeliefs, worry, and avoidance behaviors (e.g. eating disorders\citep{konstantellou2011}) may similarly be well-described by the current account. As such, the present results are transdiagnostic and not limited to one particular diagnosis. Finally, the relationship between evaluation, planning, and neural replay discussed above suggests potential future work that might help to bring this model into contact with the more memory-related aspects of PTSD. For instance, the same replay processes that can be used to evaluate actions can also, in theory, update predictive representations, cognitive maps, or models of the environment \citep{russek_predictive_2017}, such as the successor representation \cite{dayan1992,momennejad_successor_2017}. If so, similar biases in replay could result in not just abberant avoidance behavior, but also progressive, abberant remodeling of world models or cognitive maps, an observation which may connect to the rich and complex set of issues on memory involvement in PTSD.

%Note that in the present work, pessimism leads to distorted policy. Future work is required to show how, first, such pessimism can also bias offline replay with negative prediction errors (PE) rather than prioritizing surprising events based on their unsigned PE \citep{momennejad_offline_2018}.

%Recent research using human experiments and reinforcement learning models has shown that offline replay can update predictive representations of the environment \citep{momennejad_successor_2017; russek_predictive_2017}. Second, and sometimes as a result of biased replay, pessimism can be encoded in state representations or cognitive maps, e.g., transition probabilities, or multi-step successor representations \citep{momennejad_successor_2017}, that the agent uses to make decisions. Maladaptive effects of pessimistic bias on replay and predictive representations of the environment may underlie, and explain, behavioral symptoms of memory-related anxiety disorders such as post traumatic stress disorder (PTSD).

>>>>

Though anxiety disorders differ in their particular symptomology, and in the content and situations that elicit symptoms, they all are similarly characterized by aberrations in the processing of and response to threat \citep{dsm5}. In particular, at least three symptoms manifest across the anxiety disorders. First, anxiety is associated with exaggerated threat appraisal, or a bias towards evaluating threat as disproportionately greater in likelihood and severity than is warranted \citep{ClarkBeck2011}. Second, anxiety is also associated with fear generalization, wherein the primary threat becomes associated with increasingly distal locations, events, and thoughts \citep{dymond2015}. Finally, anxiety is associated with persistent avoidance behavior, which often occurs well in advance of the materialization of actual threat \citep{Arnaudova2017}. Excessive avoidance behaviors are an especially harmful aspect of anxiety disorders, both because they interfere with daily life and because they indirectly maintain anxiety by preventing learning from the non-occurrence of perceived threats. Though laboratory studies of decision making and learning in anxious populations have corroborated these clinical observations \citep{Harle2017, norbury2018, Aylward2019}, none offer an explanation as to the root of these symptoms.

@@

Although the return-maximizing assumption self-consistently defines optimal behavior, an agent need not be restricted to it \citep{symmonds2010}, and might in principle anticipate encountering danger under different (e.g., pathological) assumptions about the future. For example, an agent may expect to fail to take the correct protective actions in later states (i.e., to use a suboptimal $\pi(a \mid s)$), or may believe the world's future dynamics do not guarantee reliable avoidance even so (i.e., under stochastic or adversarial transition dynamics $P(s' \mid s,a)$). Consider an agent who has such pessimistic expectations about dangerous events at future steps. Note that assumptions of this sort, even if incorrect, can serve adaptive purposes. In general, pessimistic assumptions can help to ensure robustness and safety under uncertain or even adversarial scenarios \cite{Garcia2015}. Related work in RL shows how computing returns with pessimistic predictions can help to quantify variability in outcomes (i.e., to learn different points in the distribution of possible returns)\cite{bellemare2017}, which is one way to explicitly and tunably take account of risk tolerance. Indeed, it is common in machine learning theory to optimize outcomes under worst-case assumptions.

@@

The parameter $w$ controls the degree of pessimism. An optimistic agent ($w = 1$) expects in the future to act fully in accordance with its preferences, whereas a pessimistic agent ($w = 0$) expects to act contrary to its preferences. Importantly, this belief is false under the model: that is, we assume, at each step, that the agent actually chooses by maximizing the values given by Equation~\ref{eq:pessimism}. But these values are always computed under the assumption that the agent will then not maximize at all later steps. (Another way to say this is that we assume choices at each step are optimal, but only under the assumption that later choices will not be.) Fig.~1c,d illustrates the the consequences of different levels of pessimism for valuation in an example gridworld. The value of threat propagates, with increasing distance across the state space, due to an increasing expectation that the agent may fail to avoid threat in the future.

@@

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/02_appavo.png}}
  }
  \par \textbf{Figure 2:} (A) The Balloon Analog Risk Task (BART)\citep{Lejuez2002,Maner2007, ramirez2015}. The risk of balloon burst (point loss) increases with each pump. The optimal policy (number of pumps) under increasingly pessimistic valuation is presented for low (B) and high risk (C) balloons. The optimistic agent ($w=1$) prefers a policy reflecting the true environmental risk. The moderate ($w=0.6$) and strongly ($w=0.2$) pessimistic agents cash-out earlier, as is observed in anxious individuals. (D/E) In the sleeping predator task, the risk of loss is constant but the cost of loss still increases as more rewards are gathered. The value of reward pursuit under increasingly pessimistic valuation is presented for scenarios with low (D) and high risk (E)  of predator awakening. The relative value of approach (vs. avoid) decreases with loss amount and threat level, and moreso under pessimistic assumptions. (Parameters: $\gamma = 1.0$)
\end{figure}

@@

@@

So far, we have considered only the asymptotic preferences implied by our pessimistic value function, which we computed directly through value iteration. But we can also consider the process of learning under this value function (e.g., by variants of Q-learning\citep{SuttonBarto2018} or DYNA\citep{sutton1991, russek_predictive_2017} using the $\beta$-pessimistic return). The dynamics of such learning may speak to the progression of symptoms.

@@

\begin{figure}[!b]
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/05_choice.png}}%
  }
  \par \textbf{Figure 5:} (A) The free choice premium paradigm\cite{Leotti2011,Leotti2014} with equal chance of outcome $R \in [1, -1]$. Non-anxious participants exhibit a preference for the free choice option (blue) despite it conferring no benefit over the fixed choice option (grey). (B) Pessimistic agents show an attenuated free choice bias. The fractional preference for the free choice option over the simulated experiment is shown for three populations of subjects with levels of pessimism $w$ (y-axis; each dot represents an individual simulated agent, and the smoothed density of free choice bias for each pessimism level is also shown.)
  (Parameters: Q-learning with $\gamma = 1.0$, and inverse temperature, $\beta$, increased from 1 to 10 over 100 episodes.)
\end{figure}

@@

For instance, one variant of the model is suggested by the observation that Eq.~1 is also computed in expectation over the anticipated future environmental dynamics $p(s' \mid s,a)$. Thus, pessimism can alternatively be encoded in this distribution: e.g., a false belief that the world's response to one's choices is unpredictable or adversarial. Because the Bellman equation for the return averages over this distribution in addition to the choice policy at each step, and because an unpredictable environment also reduces the efficacy of avoidance, either formulation can produce ultimately similar results in our simulations here. A third variant of our model arises from uncertainty about the current state $s$ of the environment. Although we have taken it as fully observed, if the world state is only partly known, then this distribution too must be averaged out in evaluating each action \citep{kaelbling1998}, and here also a pessimistic skew will propagate the expectation of danger and result in exaggerated avoidance \citep{Paulus2012}. In summary, pessimistic resolution of several different varieties of uncertainty (e.g., about future action, environmental dynamics, or environmental state) could each produce similar symptoms for analogous reasons. However, from the perspective of cognitive theories of anxiety, these represent quite different maladaptive beliefs: a key difference that may be relevant in guiding treatment (especially cognitive psychotherapies aimed at ameliorating the false beliefs) of a host of anxiety disorders.

Particularly due to the way it encompasses several such variants, our account formalizes a longstanding range of theory on the role of control in anxiety. Central to many prominent cognitive theories of anxiety in the psychiatric literature is a perceived lack of control. For example, self-efficacy theory \citep{bandura1977} and the triple vulnerabilities model \citep{barlow2002} both posit that a reduced belief in the ability to effectively respond to future threat is involved in the genesis and maintenance of clinical anxiety. In contrast, and focused less on the self, the learned helplessness theories of anxiety \cite{alloy1990} claims clinical anxiety results from an uncertain belief in the controllability of the environment, such that future threat cannot be effectively mitigated or avoided. As we note above, the present model and analysis (though simulated above in terms of self-efficacy) can accommodate either variant, and shows how they relate to one another as members of a more general family of accounts.

@@

It is important to note that the present model may not describe all anxiety disorders with equal accuracy. Indeed, our analysis of pessimistic sequential evaluation is, by definition, a model of prospective cognition. Thus, the present results are more likely to accurately describe the anxiety disorders which primarily involve aberrations in future-oriented cognitive processes, such as generalized and social anxiety disorders. Naturally, the present model can account neither for the compulsive behaviors of obsessive-compulsive disorder (OCD) nor the memory disturbances of post-traumatic stress disorder (PTSD). That said, recent clinical studies suggest that diminished perceived control is a vulnerability factor common to all anxiety disorders, including OCD and PTSD \citep{gallagher2014a, gallagher2014b}. Importantly, aspects of other psychiatric disorders that involve future-oriented misbeliefs, worry, and avoidance behaviors (e.g. eating disorders\citep{konstantellou2011}) may similarly be well-described by the current account. As such, the present results are transdiagnostic and not limited to one particular diagnosis. Finally, the relationship between evaluation, planning, and neural replay discussed above suggests potential future work that might help to bring this model into contact with the more memory-related aspects of PTSD. For instance, the same replay processes that can be used to evaluate actions can also, in theory, update predictive representations, cognitive maps, or models of the environment \citep{russek_predictive_2017}, such as the successor representation \cite{dayan1993, momennejad_successor_2017}. If so, similar biases in replay could result in not just aberrant avoidance behavior, but also progressive, aberrant remodeling of world models or cognitive maps, an observation which may connect to the rich and complex set of issues on memory involvement in PTSD.
